Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package \vskip15pt Supplementary Materials

### 2018-06-22 docs/MEE_submission/manuscript_MEE.tex
Edit MS and added some stuff

\textbf{Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package}

@@

In many different contexts, researchers make use of data presented in primary literature. In the ecology and evolution, this most notably includes comparative and meta-analyses. These techniques rely on descriptive statistics (e.g. means, standard deviations (SD), sample sizes, correlation coefficients) extracted from primary literature. As well as being presented in the text or tables of research papers, descriptive statistics are frequently presented in figures and so need to be manually extracted using digitising programs.

Although there are several tools that extract data from figures (e.g. \proglang{DataThief} \citep{DataThief}, \proglang{GraphClick} \citep{GraphClick}, \proglang{WebPlotDigitizer} \citep{WebPlotDigitizer}, see Table \ref{tab:comparison}), these tools do not cater to needs of meta-analysis for four main reasons (here we focus on meta-analysis, although many  points apply to extraction for comparative analysis). First, although meta-analysis is an important tool in consolidating the data from multiple studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending or replicating studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and aid in resolving the reproducibility crises seen in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, Parker2016}. Second, digitising programs do not allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations laborious, as information has to be added later using different software. Third, existing programs do not import sets of images for the user to systematically work through. Instead they require the user to manually import images one by one, and export data into individual files, that need to be imported and edited using different software. Finally, digitising programs typically only provide the user with calibrated \textit{x,y} coordinates from imported figures, and do not differentiate between common plot types that are used to present data. Consequently a large amount of additional data manipulation is required, that is different across plots types. For example, data are frequently presented in plots with means and standard errors or confidence intervals (Figure \ref{fig:all_extract}A), from which the user wants a mean and SD for each group presented. From \textit{x,y} coordinates, users must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to SD, according to the type of error presented.

Data extraction from figures is therefore an incredibly time-consuming process as existing software does not provide an optimized, reproducible research pipeline to facilitate data extraction and editing. Here, we present an interactive \proglang{R} package, \pkg{metaDigitise} (available at https://github.com/daniel1noble/metaDigitise), which is designed for large scale, reproducible data extraction from figures, specifically catering to the the needs of meta-analysts. To this end, we provide tools to extract data from common plot types (mean/error plots, box plots, scatter plots and histograms, see Figure \ref{fig:all_extract}). \pkg{metaDigitise} operates within the \proglang{R} environment making data extraction, analysis and export more streamlined. The necessary calculations are carried out on calibrated data immediately after extraction so that comparable summary statistics can be obtained quickly. Summary data from multiple figures is returned into a single data frame which can be can easily exported or use in downstream analysis within \proglang{R}. Completed digitisations are automatically saved for each figure, meaning users can redraw their digitisations on figures, make corrections and access calibration and proceeded data. This makes sharing figure digitisation and reproducing the work of others simple and easy, and allows meta-analyses to be updated more efficiently.

@@

\section{\pkg{metaDigitise} and Reproducibility}
The \pkg{metaDigitise} package has one main function, \fct{metaDigitise}, which interactively takes the user through the process of extracting data from figures. \fct{metaDigitise} works on a directory containing images of figures copied from primary literature, in .png, .jpg, .tiff, .pdf format, specified to \fct{metaDigitise} through the \code{dir} argument. \fct{metaDigitise} recognizes all the images in the given directory and automatically imports them one by one, allowing the user to extract the relevant information about a figure as they go. It is worth the user thinking carefully about their directory structure early on in their project. Although different directory structures may be used, we would recommend having all figures for one project in a single directory with an informative and unambiguous naming scheme (e.g. paper\_figure\_trait.png). This expedites digitisation by preventing users from having to constantly change directories and / or open new images.

The data from each completed image is automatically saved as a \code{metaDigitise} object in a separate .RDS file to a \code{caldat} directory that is created within the parent directory when first executing \fct{metaDigitise}. These files enable re-plotting and editing of images at a later point (see below). When run, \fct{metaDigitise} also identifies the images within a directory that have been previously digitised and only imports new images to process. The data of all images is then automatically integrated into the final output. This means that all figures do not need to be extracted at one time and new figures can be added to the directory as the project develops.

This directory structure allows the complete digitisation process to be reproduced at a later stage, shared with collaborators and presented as supplementary materials for a publication. As long as all the images and the caldat directory are still in one directory, \fct{metaDigitise} will be able to reproduce all figure extractions, regardless of the computer it is run on. For an analysis to be updated, new figures can simply be added to the directory and \fct{metaDigitise} run to incorporate the new data.

@@

\section{Image Processing}
Running \fct{metaDigitise} presents the user with three options; `Process new images', `Import existing data' or `Edit existing data'. Selecting `Process New Images' starts the digitisation process on images within the directory that have not previously been digitised; the other functions are discussed below.

For all plot types, \fct{metaDigitise} requires the user to calibrate the axes in the figure, by clicking on two known points on the axis in question, and entering the value of those points (Figure \ref{fig:all_extract}). \fct{metaDigitise} then calculates the value of any clicked points in terms of the figure axes. This is based on the calibration used in the \pkg{digitize} R package \citep{Poisot2011}. For mean/error and box plots, only the y-axis is calibrated (Figure \ref{fig:all_extract}A,B), assuming the x-axis is redundant. For scatter plots and histograms both axes are calibrated (Figure \ref{fig:all_extract}C,D).

@@

\subsection{Mean/Error and Box Plots} 
\fct{metaDigitise} handles mean/error and box plots in a very similar way. For each mean/box, the user is enters group names and sample sizes. If the user does not enter a sample size at the time of data extraction (if, for example, the information is not readily available) a SD is not calculated. Sample sizes can, however, be entered at a later time (see below). For mean/error plots, the user clicks on an error bar and the mean. Error bars above or below the mean can be clicked, as sometimes one is clearer than the other. \fct{metaDigitise} assumes that the error bars are symmetrical. Points are displayed where the user has clicked, with the error in a different colour to the mean (Figure \ref{fig:all_extract}A). The user also enters the type of error used in the figure: SD, standard error (SE) or 95\% confidence intervals (CI95). For box plots, the user clicks on the maximum, upper quartile, median, lower quartile and minimum. For both plot types, the user can add, edit or remove groups. Three functions, \fct{error\_to\_sd}, \fct{rqm\_to\_mean} and \fct{rqm\_to\_sd}, that convert different error types to SD, box plot data to mean and box plot data SD, respectively, are also available in the package (see supplements for further details of these conversions).

\subsection{Scatter plots}
Users can extract points from multiple groups from scatter plots. Different groups are plotted in different colours and shapes to enable them to be distinguished, with a legend at the bottom of the figure (Figure \ref{fig:all_extract}C). Mean, SD and sample size are calculated from the clicked points, for each group. Data points may overlap with each other making it impossible to know whether points have been missed. This may result in the sample size of digitised groups conflicting with what is reported in the paper. For example, in Figure \ref{fig:all_extract}C only 49 points have been clicked when the sample size is known to be 50. Hence, \pkg{metaDigitise} also provides the user with the option to input known sample sizes directly. Nonetheless, it is important to recognise the impact that overlapping points can have on summary statistics, and in particular on sampling variance.

\subsection{Histograms}
The user clicks on the top corners of each bar, which are drawn in alternating colours (Figure \ref{fig:all_extract}D). Bars are numbered to allow the the user to edit them. As with scatter plots, if the sample size from the extracted data does not match a known sample size, the user can enter an alternate sample size. The calculation of mean, SD and sample size from this data is shown in the supplements.

@@

\section{Importing and Editing Previously Digitised data}
\pkg{metaDigitise} is also able to re-import, edit and re-plot previously digitised figures. When running \fct{metaDigitise}, the user can choose to `Import existing data', which returns previously digitised data, from sinlge or all figures. Alternately, the \fct{getExtracted} function returns the data of previous digitisations, but without user interaction, allowing easier integration into larger scripts. `Edit existing data' allows the user to re-plot or edit information or digitisations that have previously be done.

\subsection{Adding Sample Sizes to Previous Digitisations}
In many cases sample sizes may not be readily available when digitising figures. This information does not need to be added at the time of digitisation. To expedite finding and adding these sample sizes at a later point, \fct{metaDigitise} has a specific edit option that allows users to enter previously omitted sample sizes. This first identifies missing sample sizes in the digitised output, re-plots the relevant figures and prompts the user to enter the sample sizes for the relevant groups in the figure, one by one.

>>>>

\textbf{Reproducible, flexible and high-throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package}

@@

In the fields of ecology and evolution, researchers make use of data presented in primary literature for comparative- and meta-analyses. These techniques rely on descriptive statistics (e.g. means, standard deviations (SD), sample sizes, correlation coefficients) extracted from primary literature. As well as being presented in the text or tables of research papers, descriptive statistics are frequently presented in figures and so need to be manually extracted using digitising programs.

Although there are several tools that extract data from figures (e.g. \proglang{DataThief} \citep{DataThief}, \proglang{GraphClick} \citep{GraphClick}, \proglang{WebPlotDigitizer} \citep{WebPlotDigitizer}, see Table \ref{tab:comparison}), these tools do not cater to needs of meta-analysts for four main reasons (here we focus on meta-analysis, although many  points apply to extraction for comparative analysis). First, although meta-analysis is an important tool in consolidating the data from multiple studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending or replicating studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and aid in resolving the reproducibility crises seen in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, Parker2016}. Second, digitising programs do not allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations laborious, as information has to be added later using different software. Third, existing programs do not import sets of images for the user to systematically work through. Instead they require the user to manually import images and export the resulting digitised data into individual files one-by-one. These data often subsequently need to be imported and edited using different software. Finally, digitising programs typically only provide the user with calibrated \textit{x,y} coordinates from imported figures, and do not differentiate between common plot types that are used to present data. Consequently, a large amount of additional data manipulation is required, that is different across plots types. For example, data are frequently presented in plots with means and standard errors or confidence intervals (Figure \ref{fig:all_extract}A), from which the user wants a mean and SD for each group presented. From \textit{x,y} coordinates, users must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to SD, according to the type of error presented.

Data extraction from figures is therefore an incredibly time-consuming process as existing software does not provide an optimized, reproducible research pipeline to facilitate data extraction and editing. Here, we present an interactive \proglang{R} package, \pkg{metaDigitise} (available at https://github.com/daniel1noble/metaDigitise), which is designed for large scale, reproducible data extraction from figures, specifically catering to the the needs of meta-analysts. To this end, we provide tools to extract data from common plot types (mean/error plots, box plots, scatter plots and histograms, see Figure \ref{fig:all_extract}). \pkg{metaDigitise} operates within the \proglang{R} environment making data extraction, analysis and export more streamlined. The necessary calculations are carried out on calibrated data immediately after extraction so that comparable summary statistics can be obtained quickly. Summary data from multiple figures is returned into a single data frame which can be can easily exported or used in downstream analysis within \proglang{R}. Completed digitisations are automatically saved for each figure, meaning users can redraw their digitisations (along with metadata) on figures, make corrections and access calibration and processed (i.e., summarised) data. This makes sharing figure digitisation and reproducing the work of others simple and easy, and allows meta-analyses to be updated more efficiently.

@@

\section{\pkg{metaDigitise} and Reproducibility}
The \pkg{metaDigitise} package has one main function, \fct{metaDigitise}, which interactively takes the user through the process of extracting data from figures. \fct{metaDigitise} works on a directory containing images of figures copied from primary literature, in .png, .jpg, .tiff, .pdf format, specified to \fct{metaDigitise} through the \code{dir} argument. \fct{metaDigitise} recognizes all the images in the given directory and automatically imports them one-by-one, allowing the user to extract the relevant information about a figure as they go. Figures can be organised in different ways for a project, but we would recommend having all figures for one project in a single directory with an informative and unambiguous naming scheme (e.g. paper\_figure\_trait.png). This expedites digitisation by preventing users from having to constantly change directories and / or open new images.

The data from each completed image is automatically saved as a \code{metaDigitise} object in a separate .RDS file to a \code{caldat} folder that is created within the parent directory when first executing \fct{metaDigitise}. These files enable re-plotting and editing of images at a later point (see below). When run, \fct{metaDigitise} also identifies the images within a directory that have been previously digitised and only imports new images to process. The data of all images is then automatically integrated into the final output. This means that all figures do not need to be extracted at one time and new figures can be added to the directory as the project develops.

The complete digitisation process can then be reproduced at a later stage, shared with collaborators and presented as supplementary materials for a publication, regardless of the computer it is run on. For an analysis to be updated, new figures can simply be added to the directory and \fct{metaDigitise} run to incorporate the new data.

@@

\section{Image Processing}
Running \fct{metaDigitise} presents the user with three options; `Process new images', `Import existing data' or `Edit existing data', which can be used during and after digitisation to execute a range of functions (see Figure \ref{fig:all_extract} – 'Editing' and 'Importing' are discussed in the next section). Selecting `Process New Images' starts the digitisation process on images within the directory that have not previously been digitised. For all plot types, \fct{metaDigitise} requires the user to calibrate the axes in the figure, by clicking on two known points on the axis in question, and entering the value of those points (Figure \ref{fig:all_extract}). \fct{metaDigitise} then calculates the value of any clicked points in terms of the figure axes. This is based on the calibration used in the \pkg{digitize} R package \citep{Poisot2011}. For mean/error and box plots, only the y-axis is calibrated (Figure \ref{fig:all_extract}), assuming the x-axis is redundant. For scatter plots and histograms both axes are calibrated (Figure \ref{fig:all_extract}).

@@

\subsection{Mean/Error and Box Plots} 
\fct{metaDigitise} handles mean/error and box plots in a very similar way. For each mean/box, the user enters group name(s) and sample size(s). If the user does not enter a sample size at the time of data extraction (if, for example, the information is not readily available) a SD is not calculated. Sample sizes can, however, be entered at a later time (see next section). For mean/error plots, the user clicks on an error bar followed by the mean. Error bars above or below the mean can be clicked, as sometimes one is clearer than the other. \fct{metaDigitise} assumes that the error bars are symmetrical. Points are displayed where the user has clicked, with the error in a different colour to the mean (Figure \ref{fig:all_extract}A). The user also enters the type of error used in the figure: SD, standard error (SE) or 95\% confidence intervals (CI95). For box plots, the user clicks on the maximum, upper quartile, median, lower quartile and minimum. For both plot types, the user can add, edit or remove groups while digitising for when finished. Three functions, \fct{error\_to\_sd}, \fct{rqm\_to\_mean} and \fct{rqm\_to\_sd}, that convert different error types to SD, box plot data to mean and box plot data SD, respectively, are also available in the package (see supplements for further details of these conversions).

\subsection{Histograms}
The user clicks on the top corners of each bar, which are drawn in alternating colours (Figure \ref{fig:all_extract}C). Bars are numbered to allow the the user to edit them. As with scatter plots, if the sample size from the extracted data does not match a known sample size, the user can enter an alternate sample size. The formulas for calculation of mean, SD and sample size are provided in the supplement.

\subsection{Scatter plots}
Users can extract points from multiple groups from scatter plots. Different groups are plotted in different colours and shapes to enable them to be distinguished, with a legend at the bottom of the figure (Figure \ref{fig:all_extract}D). Mean, SD and sample size are calculated from the clicked points, for each group. Data points may overlap with each other making it impossible to know whether points have been missed. This may result in the sample size of digitised groups conflicting with what is reported in the paper. However, users also have the option to input known sample sizes directly, if required. Nonetheless, it is important to recognise the impact that overlapping points can have on summary statistics, and in particular on sampling variance.

@@

\section{Importing and Editing Previously Digitised data}
\pkg{metaDigitise} is also able to re-import, edit and re-plot previously digitised figures. When running \fct{metaDigitise}, the user can choose to 'Import existing data', which returns previously digitised data, from a single figure or all figures. Alternately, the \fct{getExtracted} function returns the data of previous digitisations, but without user interaction, allowing easier integration into larger scripts. 'Edit existing data' allows the user to re-plot or edit information for digitisations that have previously be done.

\subsection{Adding Sample Sizes to Previous Digitisations}
In many cases sample sizes may not be readily available when digitising figures. This information does not need to be added at the time of digitisation. To expedite finding and adding these sample sizes at a later point, \fct{metaDigitise} has a specific edit option that allows users to enter previously omitted sample sizes. This first identifies missing sample sizes in the digitised output, re-plots the relevant figures and prompts the user to enter the sample sizes for the relevant groups in the figure.
